---
title: "Quantify10xData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantify10xData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Comparing many 10x datasets requires the same cell filters and summary metrics
at a shared grouping grain (typically `cDNA_ID`). Labels, score columns, and
object completeness differ by species and by how each Seurat object was
prepared, which makes ad hoc aggregation difficult to perform.
`Quantify10xData()` seeks to resolve that by downloading full Seurat objects for LabKey
sequence-analysis output file IDs and scoring according to TSV specifications 
(a "spec") of grouping, filtering, and quantification rules.

The package contains bundled specs for two label systems: human cells
annotated with CellTypist Immune High and Low, and rhesus cells annotated with
the RIRA Immune, TNK, and Myeloid models. Cross-species column alignment is not
done by renaming human TargetFields. For human data, `coerceToRIRA = TRUE`
(the default) maps CellTypist labels onto RIRA Immune/TNK/Myeloid fields and
scores the bundled rhesus spec against those mapped columns, returning a second
wide table (`countsWideRIRA`) with the same TargetField names a native rhesus
run would produce.

This vignette walks through the steps offline. Instead of calling
LabKey-related functions, we build a small synthetic table of human CellTypist cells and feed it
to the same internal helpers the public function uses,
`.ParseQuantifySpec()` and `.RunQuantifyMetricLoop()`.

## Public API

In a LabKey session, call the public function with output file IDs and a spec
(a path to a `.tsv` or a `data.frame`):

```{r api-example, eval = FALSE}
library(Rdiscvr)

result <- Quantify10xData(
  outputFileIds = c(674096L),
  spec = system.file(
    "extdata", "quantify_human_immune_spec.tsv",
    package = "Rdiscvr"
  ),
  coerceToRIRA = TRUE,
  outputPrefix = "quantify_out"
)

# result$countsWide       - native human High/Low metrics
# result$countsWideRIRA   - RIRA-shaped table (human + coerceToRIRA)
# result$failures         - deferred validation / processing failures
# result$datasetSummary   - per-outputFileId ingest counts (nCellsIngested)
```

Each ID is downloaded as a full Seurat object. Missing standard `*_UCell`
scores are computed via `RIRA::CalculateUCellScores` (with `Proliferation_UCell`
added adhoc when needed), and absent `Is_TCR_Stimulated` is predicted via
`RIRA::PredictTcellActivation`. For rhesus, `classifyTNK` (default `TRUE`)
applies a TCR-aware gamma-delta and NK overrides on `RIRA_TNK_v2.cellclass`
within the `T_NK` compartment and soft-skips when those RIRA fields are absent.
Processing continues after validation or download failures; they accumulate in
`failures` and `failureLogFile`.

Everything from here on uses a synthetic cell table so you can explore the
specs, metric types, and coerce behavior without LabKey.

## Synthetic human metadata

The simulated data below mimics a minimal amount of PRIME-Seq-style metadata and is reused
in every section. It holds 24 cells across two `cDNA_ID`s, with two High/Low
CellTypist label combinations:

```{r simulatedata}
library(Rdiscvr)
library(tibble)

set.seed(42)
n_cells <- 24L
amino_acids <- c(
  "A", "C", "D", "E", "F", "G", "H", "I", "K", "L",
  "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y"
)

RandomCdr3 <- function(n) {
  cdr3 <- vapply(
    seq_len(n),
    function(i) paste0(sample(amino_acids, 12L, replace = TRUE), collapse = ""),
    character(1)
  )
  cdr3[sample.int(n, size = 4L)] <- NA_character_
  cdr3
}

cell_table <- tibble(
  sourceOutputFileId = rep(c(101L, 102L), each = 12L),
  cDNA_ID = rep(c(1001L, 1002L), each = 12L),
  `celltypist.Immune_All_High.cellclass` = rep("T cells", n_cells),
  `celltypist.Immune_All_Low.cellclass` = c(
    rep("Tcm/Naive helper T cells", 8),
    rep("Tem/Temra cytotoxic T cells", 4),
    rep("Tcm/Naive helper T cells", 6),
    rep("Tem/Temra cytotoxic T cells", 6)
  ),
  Cytotoxicity_UCell = runif(n_cells),
  Interferon_Response_UCell = runif(n_cells),
  TandNK_Activation2_UCell = runif(n_cells),
  MHCII_UCell = runif(n_cells),
  Proliferation_UCell = runif(n_cells),
  Perforin_UCell = runif(n_cells),
  Tcell_EffectorDifferentiation = sample(1:10, n_cells, replace = TRUE),
  Is_TCR_Stimulated = sample(c(TRUE, FALSE), n_cells, replace = TRUE),
  TRA = RandomCdr3(n_cells),
  TRB = RandomCdr3(n_cells),
  PDCD1 = sample(0:1, n_cells, replace = TRUE),
  GZMB = sample(0:1, n_cells, replace = TRUE),
  KLRK1 = sample(0:1, n_cells, replace = TRUE),
  HAVCR2 = sample(0:1, n_cells, replace = TRUE),
  TIGIT = sample(0:1, n_cells, replace = TRUE),
  PRF1 = sample(0:1, n_cells, replace = TRUE),
  FOXP3 = sample(0:1, n_cells, replace = TRUE),
  IL2RA = sample(0:1, n_cells, replace = TRUE)
)

cell_table
```

UCell columns are continuous scores on \([0, 1]\); gene columns are 0/1
detection flags that stand in for counts-derived positivity; and `TRA` / `TRB`
are random 12-mer amino-acid CDR3 sequences with a handful of `NA`s so diversity
rules can treat some cells as lacking a recovered chain.

## Specifications

The bundled human and rhesus specs live under `inst/extdata/`. Each row is one
metric: a grouping grain, a cell filter, and a `QuantificationType`. There are
four types:

- **`sum`** counts the cells that match.
- **`score`** reports quantiles of a numeric `QuantificationSourceField`
  (columns named `TargetField__p05`, `__median`, and `__p95`).
- **`pct_positive`** gives the percent of matched cells with a gene detected.
- **`diversity`** computes a TCR clone Hill number (q = 2) via rarefaction.

The columns that drive matching are `GroupingVariable`, `TargetField`,
`SourceField`, `MatchValue`, and `QuantificationType`. Rows for `score` and
`pct_positive` also set `QuantificationSourceField` (and, for scores,
`QuantificationScoreType = quantiles`). Optional `ScopeField` and
`ScopeMatchValue` restrict the population before the primary filter. Optional
effector-differentiation columns
(`EffectorDifferentiationScoreField`, cutpoints, and
`SubsetPhenotypeOutputFieldName`) further bin matched cells by a numeric EDS
score; both bundled specs use those bins for selected T-cell subsets.

As an example, here are two rows of each type from the human spec:

```{r spec-examples}
human_spec_path <- system.file(
  "extdata", "quantify_human_immune_spec.tsv",
  package = "Rdiscvr"
)
rhesus_spec_path <- system.file(
  "extdata", "quantify_rhesus_spec.tsv",
  package = "Rdiscvr"
)

human_spec <- Rdiscvr:::.ParseQuantifySpec(human_spec_path)$spec

spec_columns <- c(
  "GroupingVariable", "TargetField", "SourceField", "MatchValue",
  "QuantificationType", "QuantificationSourceField",
  "QuantificationScoreType", "ScopeField", "ScopeMatchValue"
)

type_order <- c("sum", "score", "pct_positive", "diversity")
spec_examples <- do.call(
  rbind,
  lapply(type_order, function(qtype) {
    rows <- human_spec[human_spec$QuantificationType == qtype, spec_columns, drop = FALSE]
    rows[seq_len(min(2L, nrow(rows))), , drop = FALSE]
  })
)
rownames(spec_examples) <- NULL
spec_examples
```

## Output

We now run the native human metric loop over the fixture. The full result is
very wide (hundreds of TargetFields), so we pull a curated slice: the grouping
key plus one column from each of the four QuantificationTypes.

```{r output}
native <- Rdiscvr:::.RunQuantifyMetricLoop(
  spec_table = human_spec,
  cell_table = cell_table,
  failures = Rdiscvr:::.EmptyFailuresTibble()
)

slice_columns <- c(
  "cDNA_ID",
  "ImmuneHigh__T_cells",
  "ImmuneLow__Tcm_Naive_helper_T_cells",
  "ImmuneHigh__T_cells__CytotoxicityScore__median",
  "ImmuneLow__Tem_Temra_cytotoxic_T_cells__PctPositive__PDCD1",
  "ImmuneLow__Tem_Temra_cytotoxic_T_cells__Diversity"
)

native$countsWide[, slice_columns, drop = FALSE]
```

A real `Quantify10xData()` call differs mainly in how the cell table is built:
it downloads each Seurat object from LabKey, computes missing `*_UCell` scores
via `RIRA::CalculateUCellScores` (with `Proliferation_UCell` added adhoc when
needed) and predicts absent `Is_TCR_Stimulated` via `RIRA::PredictTcellActivation`,
and, for rhesus, optionally reclassifies TNK labels before extraction. The metric
loop itself is the same path exercised above.

## coerceToRIRA

Cross-species alignment maps human CellTypist High/Low labels onto RIRA
Immune, TNK, and Myeloid fields using a bundled role map (Low preferred, High
as fallback, else `Unknown`), then scores the rhesus spec on those mapped
columns. The role map looks like this:

```{r coerce-artifacts}
role_map <- LoadQuantifyRoleMap()
head(role_map)
```

Native human TargetFields therefore carry `ImmuneHigh__` / `ImmuneLow__`
prefixes, while the coerced output reuses rhesus names such as `Immune__T_NK`
and `TNK__CD4plus_T_Cells`.

A few mapping details are critical. In the role map,
`RIRA_Immune_v2.cellclass == T_NK` is RIRA's mixed lymphoid compartment
(T cells, NK cells, and helper ILCs together), not "T cells only." Finer TNK
labels drive the coerced TNK metrics: helper ILCs land in `Other` and NK cells
in `NK Cells`. The `lineageRole` column is documentation only; it is not
applied at coerce time.

We can watch the mapping on the simulated dataset:

```{r coerce-map}
mapped <- Rdiscvr:::.ApplyHumanToRhesusRoleMap(cell_table, role_map)
table(mapped$`RIRA_TNK_v2.cellclass`)
```

Then run the rhesus spec on those mapped cells and compare the first few column
names against the native run:

```{r coerce-loops}
coerced <- Rdiscvr:::.RunQuantifyMetricLoop(
  spec_table = Rdiscvr:::.ParseQuantifySpec(rhesus_spec_path)$spec,
  cell_table = mapped,
  failures = Rdiscvr:::.EmptyFailuresTibble()
)

names(native$countsWide)[1:5]
names(coerced$countsWide)[1:5]
```

The contrast is the point of coerce: native output stays in CellTypist High/Low
space, while the coerced table occupies the same TargetField namespace a rhesus
dataset would produce (`Immune__`, `TNK__`, and `Myeloid__`). That shared column
space is what allows human and rhesus wide tables to be joined without renaming
rules per study.

When you pass an `outputPrefix` to `Quantify10xData()`, it writes:

- `{prefix}_counts_wide.tsv`, the "species native" human or rhesus table
- `{prefix}_counts_wide_coercedToRIRA.tsv`,  human RIRA-shaped table
  (`countsWideRIRA`)

Rhesus runs ignore `coerceToRIRA` and, when given an `outputPrefix`, write only
the native file.
