---
title: "Quantify10xData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantify10xData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`Quantify10xData()` turns per-cell metadata into grouped summary metrics. You
give it one or more LabKey sequence-analysis output file IDs, and it downloads
the per-cell metadata for each of those files and then computes whatever
metrics you have described in a tab-separated specification. The package ships
with ready-made specification documents (so called specs) for two common situations: human cells labeled with
CellTypist Immune (at both the High and Low levels of granularity) and rhesus
cells labeled with the RIRA Immune, TNK, and Myeloid models.

This vignette walks through the process offline. Instead of talking to
LabKey, we build a small synthetic table of human CellTypist cells and feed it
to the same internal helpers the public function uses under the hood,
`.ParseQuantifySpec()` and `.RunQuantifyMetricLoop()`.

One detail is worth flagging up front. When you run human data, the function
can also emit a second, RIRA-shaped wide table. This is controlled by
`coerceToRIRA`, which is `TRUE` by default. The idea is that human cells are
first mapped onto RIRA's Immune/TNK/Myeloid bins, and the bundled rhesus spec
is then scored against those mapped columns. We come back to this at the end.

## Public API

In a real LabKey session you would simply call the public function with your
output file IDs and a spec (either a path to a `.tsv` or a `data.frame`):

```{r api-example, eval = FALSE}
library(Rdiscvr)

result <- Quantify10xData(
  outputFileIds = c(674096L),
  spec = system.file(
    "extdata", "quantify_human_immune_spec.tsv",
    package = "Rdiscvr"
  ),
  coerceToRIRA = TRUE,
  outputPrefix = "quantify_out"
)

# result$countsWide       - native human High/Low metrics
# result$countsWideRIRA   - RIRA-shaped table (human + coerceToRIRA)
# result$failures         - deferred validation / processing failures
# result$laneSummary      - per-lane ingest counts
```

Everything from here on uses a synthetic cell table so you can explore the
specs, the different metric types, and the coerce behavior without ever
touching LabKey.

## Synthetic human metadata

We start by simulating data, similar to populated metadata from PRIME-Seq and reuse it for every section
below. It holds roughly 24 cells spread across two `cDNA_ID`s, with two
combinations of High/Low CellTypist labels:

```{r fixture}
library(Rdiscvr)
library(tibble)

set.seed(42)
n_cells <- 24L
amino_acids <- c(
  "A", "C", "D", "E", "F", "G", "H", "I", "K", "L",
  "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y"
)

RandomCdr3 <- function(n) {
  cdr3 <- vapply(
    seq_len(n),
    function(i) paste0(sample(amino_acids, 12L, replace = TRUE), collapse = ""),
    character(1)
  )
  cdr3[sample.int(n, size = 4L)] <- NA_character_
  cdr3
}

cell_table <- tibble(
  sourceOutputFileId = rep(c(101L, 102L), each = 12L),
  cDNA_ID = rep(c(1001L, 1002L), each = 12L),
  `celltypist.Immune_All_High.cellclass` = rep("T cells", n_cells),
  `celltypist.Immune_All_Low.cellclass` = c(
    rep("Tcm/Naive helper T cells", 8),
    rep("Tem/Temra cytotoxic T cells", 4),
    rep("Tcm/Naive helper T cells", 6),
    rep("Tem/Temra cytotoxic T cells", 6)
  ),
  Cytotoxicity_UCell = runif(n_cells),
  Interferon_Response_UCell = runif(n_cells),
  TandNK_Activation2_UCell = runif(n_cells),
  MHCII_UCell = runif(n_cells),
  Proliferation_UCell = runif(n_cells),
  Perforin_UCell = runif(n_cells),
  Tcell_EffectorDifferentiation = sample(1:10, n_cells, replace = TRUE),
  Is_TCR_Stimulated = sample(c(TRUE, FALSE), n_cells, replace = TRUE),
  TRA = RandomCdr3(n_cells),
  TRB = RandomCdr3(n_cells),
  PDCD1 = sample(0:1, n_cells, replace = TRUE),
  GZMB = sample(0:1, n_cells, replace = TRUE),
  KLRK1 = sample(0:1, n_cells, replace = TRUE),
  HAVCR2 = sample(0:1, n_cells, replace = TRUE),
  TIGIT = sample(0:1, n_cells, replace = TRUE),
  PRF1 = sample(0:1, n_cells, replace = TRUE),
  FOXP3 = sample(0:1, n_cells, replace = TRUE),
  IL2RA = sample(0:1, n_cells, replace = TRUE)
)

cell_table
```

A few notes on the columns so the numbers below make sense. The UCell columns
are sampled uniformly on \([0, 1]\), the gene columns are simple 0/1 detection
flags, and `TRA` / `TRB` are random 12-mer amino-acid CDR3 sequences with a
handful of `NA`s mixed in to stand in for cells with no recovered chain.

## Specifications

The bundled human and rhesus specs live under `inst/extdata/`. A spec is just a
table, and each row describes one metric: how to group the cells, how to filter
them, and what kind of `QuantificationType` to compute. There are four types:

- **`sum`** counts the cells that match.
- **`score`** reports quantiles of a numeric `QuantificationSourceField`.
- **`pct_positive`** gives the percent of matched cells with a gene detected.
- **`diversity`** computes a TCR clone Hill number (q = 2) via rarefaction.

The columns that do most of the work are `GroupingVariable`, `TargetField`,
`SourceField`, `MatchValue`, and `QuantificationType`. Rows for `score` and
`pct_positive` additionally set `QuantificationSourceField` (and, for scores,
`QuantificationScoreType = quantiles`). Finally, the optional `ScopeField` and
`ScopeMatchValue` let you narrow the cell population before the primary filter
is applied.

As an example, here are a couple of example rows of each type, pulled from the human spec:

```{r spec-examples}
human_spec_path <- system.file(
  "extdata", "quantify_human_immune_spec.tsv",
  package = "Rdiscvr"
)
rhesus_spec_path <- system.file(
  "extdata", "quantify_rhesus_spec.tsv",
  package = "Rdiscvr"
)

human_spec <- Rdiscvr:::.ParseQuantifySpec(human_spec_path)$spec

spec_columns <- c(
  "GroupingVariable", "TargetField", "SourceField", "MatchValue",
  "QuantificationType", "QuantificationSourceField",
  "QuantificationScoreType", "ScopeField", "ScopeMatchValue"
)

type_order <- c("sum", "score", "pct_positive", "diversity")
spec_examples <- do.call(
  rbind,
  lapply(type_order, function(qtype) {
    rows <- human_spec[human_spec$QuantificationType == qtype, spec_columns, drop = FALSE]
    rows[seq_len(min(2L, nrow(rows))), , drop = FALSE]
  })
)
rownames(spec_examples) <- NULL
spec_examples
```

## Output

Now we run the native human metric loop over the fixture. The full result is
very wide (hundreds of TargetFields), so rather than print everything we pull
out a curated slice: the grouping key plus one column from each of the four
QuantificationTypes.

```{r output}
native <- Rdiscvr:::.RunQuantifyMetricLoop(
  spec_table = human_spec,
  cell_table = cell_table,
  failures = Rdiscvr:::.EmptyFailuresTibble()
)

slice_columns <- c(
  "cDNA_ID",
  "ImmuneHigh__T_cells",
  "ImmuneLow__Tcm_Naive_helper_T_cells",
  "ImmuneHigh__T_cells__CytotoxicityScore__median",
  "ImmuneLow__Tem_Temra_cytotoxic_T_cells__PctPositive__PDCD1",
  "ImmuneLow__Tem_Temra_cytotoxic_T_cells__Diversity"
)

native$countsWide[, slice_columns, drop = FALSE]
```
The only major between this vignette and a proper invocation is that `Quantify10xData()` first downloads the Seurat Object from
LabKey, rather than reading it from a simulated tibble.

## coerceToRIRA

Aligning columns across species does works by mapping onto RIRA's Immune, TNK, and
Myeloid labels using a bundled role map, and the rhesus spec is then scored on
those mapped columns. Let's look at the role map itself:

```{r coerce-artifacts}
role_map <- LoadQuantifyRoleMap()
head(role_map)
```

So the native human TargetFields carry `ImmuneHigh__` / `ImmuneLow__` prefixes,
while the coerced output reuses the rhesus names such as `Immune__T_NK` and
`TNK__CD4plus_T_Cells`.

A couple of subtleties are worth calling out here. In the role map,
`RIRA_Immune_v2.cellclass == T_NK` refers to RIRA's mixed lymphoid compartment
(T cells, NK cells, and helper ILCs together), not "T cells only." The finer
TNK labels are what actually drive the coerced TNK metrics, so, for example,
helper ILCs land in `Other` and NK cells in `NK Cells`. And the `lineageRole`
column is there purely as documentation; it is not applied at coerce time.

We can watch the mapping happen on the fixture:

```{r coerce-map}
mapped <- Rdiscvr:::.ApplyHumanToRhesusRoleMap(cell_table, role_map)
table(mapped$`RIRA_TNK_v2.cellclass`)
```

And then run the rhesus spec on those mapped cells, comparing the first few
column names against the native run:

```{r coerce-loops}
coerced <- Rdiscvr:::.RunQuantifyMetricLoop(
  spec_table = Rdiscvr:::.ParseQuantifySpec(rhesus_spec_path)$spec,
  cell_table = mapped,
  failures = Rdiscvr:::.EmptyFailuresTibble()
)

names(native$countsWide)[1:5]
names(coerced$countsWide)[1:5]
```

The contrast is the primary point: the native output carries High/Low columns,
whereas the coerced output lines up with the column names that a rhesus dataset
would produce (the `Immune__`, `TNK__`, and `Myeloid__` prefixes).

Finally, when you pass an `outputPrefix` to `Quantify10xData()`, it writes the
tables to disk:

- `{prefix}_counts_wide.tsv` holds the native human (or rhesus) table.
- `{prefix}_counts_wide_coercedToRIRA.tsv` holds the human RIRA-shaped table.

Rhesus runs ignore `coerceToRIRA` entirely and, when given an `outputPrefix`,
write only the native file.
